---
title: "Getting Started with switchbox"
vignette: >
  %\VignetteIndexEntry{Getting Started with switchbox}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
bibliography: switchBox.bib
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
library(switchbox)
```

## Introduction

The `switchbox` package implements the K-Top-Scoring-Pair (KTSP) algorithm for
binary classification based on pairwise feature comparisons
[@afsari2015switchbox; @marchionni2013simple]. The classifier uses the relative
ordering of feature values (e.g. gene expression levels) to make predictions,
making it:
- **Robust** to data normalization and batch effects
- **Transparent** --- decision rules are simple comparisons (gene A > gene B)
- **Compact** --- a small number of feature pairs suffice for classification

This vignette demonstrates the core workflow: training a KTSP classifier,
classifying new samples, evaluating performance, and creating diagnostic plots.

## Example Data

The package ships with breast cancer expression data from the MammaPrint
study. The training set has 70 genes measured across 78 samples, and the
test set has 307 samples.

```{r}
#| label: load-data
data(trainingData)
data(testingData)

dim(matTraining)
table(trainingGroup)

dim(matTesting)
table(testingGroup)
```

## Training a KTSP Classifier

Use `swap_train_ktsp()` to train a classifier. By default it:

1. Filters features using the Wilcoxon rank-sum test (`swap_filter_wilcoxon()`)
2. Computes signed TSP scores via C-backed routines
3. Selects the number of pairs K using a t-test criterion (`swap_k_by_ttest()`)

```{r}
#| label: train-default
classifier <- swap_train_ktsp(matTraining, trainingGroup, k_range = 3:8)
classifier$name
classifier$TSPs
```

The classifier contains `r nrow(classifier$TSPs)` top-scoring pairs. Each row
of `classifier$TSPs` is a pair (gene1, gene2) where gene1 < gene2 is expected
in one class.

### Training a Single-TSP Classifier

For the simplest possible classifier (one pair):

```{r}
#| label: train-1tsp
classifier_1tsp <- swap_train_1tsp(matTraining, trainingGroup)
classifier_1tsp$TSPs
```

### Customizing the Feature Filter

You can skip filtering by setting `filter_fn = NULL`, or provide a custom
filter:

```{r}
#| label: train-nofilter
classifier_nf <- swap_train_ktsp(
  matTraining, trainingGroup,
  filter_fn = NULL,
  k_range = 3:5
)
classifier_nf$name
```

### Restricted Pairs

If prior knowledge suggests specific candidate pairs, pass them as a
two-column character matrix:

```{r}
#| label: train-restricted
genes <- rownames(matTraining)
restricted <- cbind(
  rep(genes[1:10], each = 5),
  rep(genes[11:15], times = 10)
)
classifier_rp <- swap_train_ktsp(
  matTraining, trainingGroup,
  restricted_pairs = restricted,
  k_range = 2:5
)
classifier_rp$TSPs
```

### Handling Ties

Enable tie-aware scoring with `handle_ties = TRUE`:

```{r}
#| label: train-ties
classifier_ties <- swap_train_ktsp(
  matTraining, trainingGroup,
  handle_ties = TRUE,
  k_range = 3:6
)
classifier_ties$tie_vote
```

### Selecting K by Performance Metric

Instead of the default t-test, select K by maximizing AUC or accuracy on
training data:

```{r}
#| label: train-kbymeasurement
classifier_auc <- swap_train_ktsp(
  matTraining, trainingGroup,
  k_range = 2:8,
  k_selection_fn = swap_k_by_measurement,
  k_opts = list(measurement = "auc")
)
classifier_auc$name
```

## Classification

### Computing Voting Statistics

`swap_ktsp_statistics()` returns per-sample aggregate statistics and a
matrix of individual pair comparisons:

```{r}
#| label: statistics
stats <- swap_ktsp_statistics(matTesting, classifier)
head(stats$statistics)
stats$comparisons[1:5, ]
```

### Predicting Class Labels

`swap_ktsp_classify()` returns predicted class labels via majority voting:

```{r}
#| label: classify
predictions <- swap_ktsp_classify(matTesting, classifier)
table(predictions, testingGroup)
```

## Performance Evaluation

### Prediction Statistics

```{r}
#| label: pred-stats
perf <- swap_prediction_stats(predictions, testingGroup)
perf
```

### Combined Train-and-Test Evaluation

`swap_train_test_results()` trains a classifier and evaluates it on both
sets in one call:

```{r}
#| label: train-test
result <- swap_train_test_results(
  matTraining, trainingGroup,
  matTesting, testingGroup,
  k_range = 3:8,
  decision_values = TRUE
)
result$train
result$test
```

## Cross-Validation

### K-Fold Cross-Validation

```{r}
#| label: cv
set.seed(42)
cv_result <- swap_ktsp_cv(
  matTraining, trainingGroup,
  k = 5,
  k_range = 3:6
)
cv_result$stats
```

### Leave-One-Out Cross-Validation

LOO is more computationally intensive but avoids random fold assignment.
Here we use a subset for speed:

```{r}
#| label: loo
small_x <- matTraining[1:30, 1:30]
small_y <- trainingGroup[1:30]
loo_result <- swap_ktsp_loo(small_x, small_y, k_range = 2:4)
loo_result$stats
```

## Visualization

The package provides several ggplot2-based diagnostic plots.

### Gene Expression Across Samples

```{r}
#| label: plot-genes
#| fig-width: 8
#| fig-height: 4
swap_plot_genes(
  matTraining, trainingGroup,
  genes = classifier$TSPs[1, ]
)
```

### Gene Pair Scatter Plot

```{r}
#| label: plot-scatter
#| fig-width: 5
#| fig-height: 5
swap_plot_gene_pair_scatter(
  matTraining, trainingGroup,
  genes = classifier$TSPs[1, ]
)
```

### ROC Curves

```{r}
#| label: plot-roc
#| fig-width: 5
#| fig-height: 5
swap_plot_roc(result)
```

### Vote Heatmap

```{r}
#| label: plot-votes
#| fig-width: 10
#| fig-height: 4
swap_plot_votes(classifier, matTraining, trainingGroup)
```

### Gene Pair Boxplots

```{r}
#| label: plot-boxplot
#| fig-width: 5
#| fig-height: 4
swap_plot_gene_pair_boxplot(
  classifier$TSPs[1, ],
  matTraining, trainingGroup,
  points = TRUE,
  point_coloring = "byClass"
)
```

### Boxplots by Class

```{r}
#| label: plot-classes-boxplot
#| fig-width: 6
#| fig-height: 4
swap_plot_gene_pair_classes_boxplot(
  classifier$TSPs[1, ],
  matTraining, trainingGroup,
  points = TRUE
)
```

## Score Computation

For advanced users, the scoring infrastructure is directly accessible:

```{r}
#| label: scores
scores <- swap_calculate_scores(matTraining, trainingGroup)
head(sort(abs(scores$score), decreasing = TRUE))
```

Build a custom TSP table:

```{r}
#| label: tsp-table
tbl <- swap_make_tsp_table(scores, maxk = 10, disjoint = TRUE)
head(tbl)
```

## Utility Functions

### Train/Test Splitting

```{r}
#| label: split
set.seed(123)
split <- swap_make_train_test_data(matTraining, trainingGroup, p = 0.7)
dim(split$train_x)
dim(split$test_x)
```

### Score Format Conversion

```{r}
#| label: score-convert
m <- matrix(1:9, 3, 3, dimnames = list(c("a", "b", "c"), c("a", "b", "c")))
v <- swap_score_matrix_to_vector(m)
v[1:4]
```

## Session Info

```{r}
#| label: session-info
sessionInfo()
```

## References
