---
title: "Using switchbox with tidymodels"
vignette: >
  %\VignetteIndexEntry{Using switchbox with tidymodels}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

```{r}
#| label: setup
#| include: false
library(switchbox)
library(rsample)
library(yardstick)
library(dplyr)
library(purrr)
```

## Introduction

The tidymodels ecosystem provides a consistent interface for resampling
(`rsample`), performance metrics (`yardstick`), and data manipulation
(`dplyr`, `purrr`). This vignette shows how to use these packages alongside
switchbox for cross-validated model evaluation.

The main challenge is that switchbox expects **features in rows** and
**samples in columns**, while tidymodels works with **samples in rows** and
**features in columns**. The helper functions `swap_from_tidy()`,
`swap_to_tidy()`, and `swap_tidy_result()` bridge this gap.

## Data Format Conversion

### From tidy to switchbox

Suppose you have data in a tidy data frame with one row per sample:

```{r}
#| label: from-tidy-demo
df <- data.frame(
  outcome = factor(c("Good", "Bad", "Good", "Bad", "Good")),
  gene_A = c(2.1, 5.3, 1.8, 4.9, 2.5),
  gene_B = c(7.2, 3.1, 6.8, 2.9, 7.0)
)
df
```

Convert to switchbox format with `swap_from_tidy()`. When you specify an
`outcome` column, it returns both the matrix and the labels:

```{r}
#| label: from-tidy-convert
result <- swap_from_tidy(df, outcome = "outcome")
result$x
result$y
```

### From switchbox to tidy

The reverse operation uses `swap_to_tidy()`:

```{r}
#| label: to-tidy-demo
data(trainingData)
tidy_df <- swap_to_tidy(matTraining, trainingGroup, outcome = "prognosis")
tidy_df[1:5, 1:5]
```

## Train/Test Split with rsample

Using `rsample::initial_split()` with the package's built-in breast cancer
data:

```{r}
#| label: rsample-split
data(trainingData)

# Create a tidy data frame for rsample
full_df <- swap_to_tidy(matTraining, trainingGroup, outcome = "prognosis")

set.seed(42)
split <- initial_split(full_df, prop = 0.7, strata = prognosis)

train_df <- training(split)
test_df <- testing(split)

cat("Training:", nrow(train_df), "samples\n")
cat("Testing: ", nrow(test_df), "samples\n")
```

Convert back to switchbox format, train, and evaluate:

```{r}
#| label: rsample-train-eval
train_data <- swap_from_tidy(train_df, outcome = "prognosis")
test_data <- swap_from_tidy(test_df, outcome = "prognosis")

classifier <- swap_train_ktsp(
  train_data$x, train_data$y,
  k_range = 3:6
)

# Classify test samples and get yardstick-compatible results
results <- swap_tidy_result(classifier, test_data$x, test_data$y)
head(results)
```

Evaluate with yardstick:

```{r}
#| label: rsample-metrics
metrics <- metric_set(accuracy, sensitivity, specificity, bal_accuracy)
metrics(results, truth = truth, estimate = .pred_class)
```

## Cross-Validation with rsample

### Setting Up Folds

```{r}
#| label: cv-setup
set.seed(123)
folds <- vfold_cv(full_df, v = 5, strata = prognosis)
folds
```

### Per-Fold Training and Evaluation

Define a function that trains a switchbox classifier on the analysis set
(training fold) and evaluates it on the assessment set (validation fold):

```{r}
#| label: cv-function
evaluate_fold <- function(split) {
  train_df <- analysis(split)
  test_df <- assessment(split)

  train_data <- swap_from_tidy(train_df, outcome = "prognosis")
  test_data <- swap_from_tidy(test_df, outcome = "prognosis")

  classifier <- swap_train_ktsp(
    train_data$x, train_data$y,
    k_range = 3:6
  )

  swap_tidy_result(classifier, test_data$x, test_data$y)
}
```

Run across all folds:

```{r}
#| label: cv-run
fold_results <- map(folds$splits, evaluate_fold)
```

### Computing Per-Fold Metrics

```{r}
#| label: cv-metrics
metrics_fn <- metric_set(accuracy, sensitivity, specificity, bal_accuracy)

fold_metrics <- map_dfr(fold_results, function(res) {
  metrics_fn(res, truth = truth, estimate = .pred_class)
}, .id = "fold")

fold_metrics
```

### Aggregating Across Folds

Compute mean and standard error for each metric:

```{r}
#| label: cv-summary
cv_summary <- fold_metrics |>
  group_by(.metric) |>
  summarize(
    mean = mean(.estimate),
    se = sd(.estimate) / sqrt(n()),
    .groups = "drop"
  )
cv_summary
```

### Confusion Matrix

Pool predictions from all folds for an overall confusion matrix:

```{r}
#| label: cv-confusion
all_results <- bind_rows(fold_results)
conf_mat(all_results, truth = truth, estimate = .pred_class)
```

## Comparison with Built-In Cross-Validation

The switchbox package includes `swap_ktsp_cv()` for convenience. Here is
the same 5-fold CV for comparison:

```{r}
#| label: builtin-cv
set.seed(123)
builtin_cv <- swap_ktsp_cv(
  matTraining, trainingGroup,
  k = 5,
  k_range = 3:6
)
builtin_cv$stats
```

The tidymodels approach gives you more flexibility: stratified folds,
custom metric sets, and easy integration with other tidymodels components.
The built-in function is simpler when you just need a quick estimate.

## Session Info

```{r}
#| label: session-info
sessionInfo()
```
